LabMate: a prospectus for types for MATLAB

### 2024-04-16 writing/IMEKOWC2024/LabMate/paper.tex
[ writing ] fixes and suggestions by NPL

\title{LabMate: a prospectus for types for MATLAB}
\author{Conor McBride  \affiliationmark{1}, Georgi Nakov  \affiliationmark{1}, Fredrik Nordvall Forsberg  \affiliationmark{1}, Andr\'e Videla  \affiliationmark{1}, Alistair Forbes  \affiliationmark{2}, Keith Lines  \affiliationmark{2}}
\affiliation{%
\affiliationmark{1}{\fontsize{11pt}{9pt} \selectfont University of Strathclyde , UK}  \\
\affiliationmark{2}{\fontsize{11pt}{9pt} \selectfont National Physical Laboratory, UK}}

\maketitle

\begin{abstract}
  Many computations in science and engineering are implemented in the
  programming language MATLAB. However the high-level meaning of such
  MATLAB programs stay informal, which can lead to implementation
  errors and bugs, for example relating to incompatible units of
  measure for quantities, or incompatible sizes of matrices at
  runtime. We are in the process of developing LabMate, which is a
  tool for reifying current informal programmer practices into a
  language of formal comments. These comments are ignored by MATLAB,
  but acted on and checked by LabMate. We outline the design
  principles behind LabMate, our current progress, and our future
  plans.
\end{abstract}

@@

To rectify this situation, we have developed LabMate, which is a tool
to reify current virtuous engineering practices as a formal language
of MATLAB comments. MATLAB remains in the dark, but LabMate reads,
assesses, and transforms MATLAB programs in accordance with these
comments.
%
Behind the scenes, LabMate is retrofitting an expressive type system
to MATLAB, with the more of the meaning of programs recorded in their
types.
%
This gives a lightweight and low-cost way for MATLAB programmers to
express their intent, and in a language they are already working in,
rather than starting over from scratch.
%
While type systems and their theory is a well established discipline
of computer science, the development of LabMate has several novel
advances on the algebraic structure required to classify
matrices and the meanings of the quantities therein, e.g., their
physical dimensions. LabMate brings advanced type systems to
effectiveness within, rather than instead of, existing scientific and
engineering toolchains and practices.

LabMate is under constant development; the latest version can be found at \url{https://github.com/msp-strath/LabMate/tree/devel}.
%
At time of writing, the necessary core type system is in place: we are busily extending the variety of MATLAB expressions it can make sense of, and the constraints it can solve.
%
We are close to covering the examples in this paper, but not quite there yet.
%
We aim to eliminate the speculative future tense from our examples before camera-ready copy is required.
%
This paper effectively documents how far we have come, and the roadmap to our next milestone.

@@

An electrical current with a known value (the applied signal) is passed through the resistor being measured, and a series of voltage readings are taken at fixed time intervals (the recorded output). After a specified number of readings are taken, the direction of the input signal must be reversed in order to separate the offset and drift from the voltage readings. Therefore, the applied signal (bridge energisation) is a square wave of {\em reversals}. A calibration factor is also applied, dividing the signal into two parts with different amplitudes.

The recorded output follows the input signal, superimposed on a detector with noise, offset and drift. Figure~1%\ref{fig:rec_out}
provides an example of recorded output, using simulated data.
\begin{figure}[H]
  \begin{center}
    \includegraphics[width=0.6\linewidth]{example_output.jpg}
    \captionwide{Simulated example of recorded output.}
    \label{fig:rec_out}
  \end{center}
\end{figure}

@@

We begin by declaring the types of the input data: the \texttt{times} and the \texttt{voltages}. Our example calculation uses twelve time and voltage inputs, arranged as two rows to be read from a file. A real measurement would involve considerably larger datasets.
\begin{verbatim}
%> times    :: [ 1 x 12 ] double
%> voltages :: [ 1 x 12 ] double
\end{verbatim}
Lines starting with `\texttt{\%>}' are interpreted as ordinary comments by MATLAB, but as \remph{directives} by LabMate, here giving us enough information not only to insist that the values of \texttt{times} and \texttt{voltages} \remph{conform} to the given sizes, but to determine \remph{how} to read them from a file. We would issue the further directive
\begin{verbatim}
%> readfrom "inputs.txt" times voltages
\end{verbatim}
LabMate is a program \remph{transducer}: its output is a modified version of its input, allowing it to respond to directives with information in comments or by generating code. Here we would expect to see something like
\begin{verbatim}
%> readfrom "inputs.txt" times voltages
%<{
h=fopen("inputs.txt");
c=textscan(h,'%f');
fclose(h);
src = c{1};
readPtr = 1;
for i = 1:12
  times[i] = src[readPtr];
  readPtr = readPtr + 1;
end
for i = 1:12
  voltages[i] = src[readPtr];
  readPtr = readPtr + 1;
end
%<}
\end{verbatim}
where the comment lines `\pct\res\lcb' and `\pct\res\rcb' visually delimit the extent of the generated boilerplate code, so that humans may safely ignore it. LabMate knows enough about scope in MATLAB to ensure that auxiliary variable naming never creates conflict. We have kept to one datum per line for simplicity, but we are free to extend our format specification language as we see fit~\cite{mgen}.

@@

Under the hood, LabMate translates selected well behaved MATLAB expressions
to typed expressions in its own core theory of types and values.
%
As matrices feature heavily in MATLAB code, the types of matrices
play a central role in this theory.
%
In particular, we acknowledge that the rows and columns of a matrix might pertain to distinct individual entities, rather as the rows ans columns of a spreadsheet do.
%
Internally, a matrix type looks like
\[\Matrix{R}{C}{E}{\rs}{\cs}
\]
%
We thus \param{} matrix types by two arbitrary `header' types $R$, for the individuals the rows concern, and $C$ for the columns. We may then give two lists, $\rs = [r_1,\ldots,r_m]\in\List{R}$ (so that each $r_i\in R$) and $\cs = [c_1,\ldots,c_n]\in\List{C}$ of those individuals. A \param{}d type $E(r,c)$ gives the type of entry for any combination of row and column individual, with $E(r_i,c_j)$ being the type of the entry $e_{i,j}$. In the diagram below, the $e$s inside the rectangle are the matrix entries computed by the MATLAB code. The $r$s and $c$s outside the rectangle are \remph{meta}data coming from the matrix type, known only to LabMate, and not available as data to MATLAB.
\[\begin{array}{@{}r|ccccc|}
\; & c_1      & \cdots & c_j     & \cdots & c_n     \\
\hline
r_1    & e_{1,1}  &        & e_{1,j} &        & e_{1,n} \\
\vdots &          &        &         &        &         \\
r_i    & e_{i,1}  &        & e_{i,j} &        & e_{i,n} \\
\vdots &          &        &         &        &         \\
r_m    & e_{m,1}  &        & e_{m,j} &        & e_{m,n}  \\
\hline
\end{array}\]
To check that $\Matrix{R}{C}{E}{\rs}{\cs}$ is a valid type, LabMate must:
\begin{enumerate}
\item check that $R$ and $C$ are both valid types;
\item check that $E(r,c)$ is a valid type whenever $r\in R$ and $c\in C$;
\item check that $\rs\in\List{R}$ and $\cs\in\List{C}$.
\end{enumerate}

@@

LabMate elaborates MATLAB expressions and commands into terms of its own internal core type theory.
%
Although the simple examples in this paper use \emph{constant} sizes and dimensions, any practicable library requires size, header and dimension expressions to contain \emph{parameters}.
%
For example, if \texttt{A} is $i\times j$ and \texttt{B} is $i\times k$, then
\begin{verbatim}
[ A B
  B A ]
\end{verbatim}
should be a valid matrix pasting, because $j+k = k+j$. Our notion of equality for numbers thus induces commutative monoid structure.
%
Physical dimensions are modelled as the free Abelian group generated by some base set of dimensions~\cite{dim.group}, so we decide that theory, too.
%
The core theory we need is in place.

@@

Coming back to the example from Section~\ref{sec:example} again, we aim to give both \texttt{M} and \texttt{voltages} more precise types, also taking their physical dimensions into account,
\begin{verbatim}
%> M :: [ 12 x i::D<-[{},{},{},{},T] ] Q(i)
%> voltages :: [ 1 x 12 ] Q(V)
\end{verbatim}
  but we had better unpack what we mean.

@@

Secondly, we see \texttt{Q(}$d$\texttt{)} as the type of \remph{quantities} with physical dimension $d$, represented in MATLAB as a mere \texttt{double} with standardised units. We shall define a type $\texttt{D}$ of dimensions presently. The notation
\begin{verbatim}
i::D<-[{},{},{},{},T]
\end{verbatim}
will tell LabMate that the column header type is $\texttt{D}$, and that the matrix entry type \texttt{Q(i)} varies in accordance with the header, locally named \texttt{i}, which is drawn from the given list of dimensions.
%
But what type is \texttt{D}? We will have defined, in LabMate directives, rather than MATLAB,
\begin{verbatim}
%{>
Base =
  ['Time, 'Length, 'Mass, 'Current,
   'Temperature, 'Amount, 'Luminosity]
D = Abel(Enum(Base))
%}
\end{verbatim}
where \texttt{Base} is a list of symbolic constants standing for base dimensions, \texttt{Enum} converts any such list to a finite enumeration type, and \texttt{Abel} constructs the free Abelian group on any type of generators, with unit \lcb\rcb, singleton generators \lcb$g$\rcb, multiplication \texttt{*}, division \texttt{/}, and exponentiation by integer constants \crt. We further define
\begin{verbatim}
%{>
T = {'Time}
V = {'Mass}*{'Length}^2/T^3/{'Current}
%}
\end{verbatim}
and \texttt{voltages} becomes a row vector uniformly containing quantities in \texttt{Q(V)}.

With this refinement in place
\begin{verbatim}
x = M \ voltages'
%> typeof x
\end{verbatim}
will yield
\begin{verbatim}
%< x :: [ i::D<-[{},{},{},{},T] ] Q(V/i)
\end{verbatim}

@@

However, if all LabMate does is complain about mistakes, it will not appeal to programmers in a hurry.
%
We have already shown a simple example where LabMate does not merely check but rather generates boilerplate code using its types as specification.
%
There is considerable potential to push further in this direction, little by little.
%
For example, we could equip LabMate with its own expression language for calculating with quantities which have not only dimensions but \remph{units}, compiling to MATLAB with appropriate conversion factors correctly inserted.
%
Our mission is, ever so gently, to tilt the balance towards programming at a higher level of structure and meaning in LabMate's directive language.

@@

Inspired again by Kennedy's work --- his gradualist reformation of PHP~\cite{hack} ---
we have set out on this journey to bring the benefits of expressive type systems to scientists and engineers \remph{where they are}.
%
It would be unrealistic of us to expect practitioners to abandon their existing toolchains and learn whole new programming styles and languages that, in any case, exist only as academic research prototypes.
%
Moreover, it has very much \remph{not} turned out that, in addressing MATLAB, we were merely \remph{applying} type system research already well established.
%
This project has pushed us in new directions and generated raised expectations of the \remph{algebra} to be done by the typechecker.
%
There is so much to learn outside our ivory tower!

@@

Thanks to NPL colleague Nick Fletcher for his help and guidance providing the example from Section \ref{sec:example}. Thanks also to NPL colleagues Louise Wright and Ian Smith for reviewing this paper, and to Professor Neil Ghani for his support.

@@

>>>>

\title{LabMate: a prospectus for types for MATLAB}
\author{Conor McBride  \affiliationmark{1}, Georgi Nakov  \affiliationmark{1}, Fredrik Nordvall Forsberg  \affiliationmark{1}, Andr\'e Videla  \affiliationmark{1}, Alistair Forbes  \affiliationmark{2}, Keith Lines  \affiliationmark{2}}
\affiliation{%
\affiliationmark{1}{\fontsize{11pt}{9pt} \selectfont University of Strathclyde , UK}  \\
\affiliationmark{2}{\fontsize{11pt}{9pt} \selectfont National Physical Laboratory, UK}} % \\
%\affiliationmark{*}{Corresponding author. Email address: \url{fredrik.nordvall-forsberg@strath.ac.uk}}}
\maketitle

\begin{abstract}
  Many computations in science and engineering are implemented in the
  programming language MATLAB. However the high-level meaning of such
  MATLAB programs stays informal, which can lead to implementation
  errors and bugs, for example relating to incompatible units of
  measure for quantities, or incompatible sizes of matrices at
  runtime. We are in the process of developing LabMate, which is a
  tool for reifying current informal programmer practices into a
  language of formal comments. These comments are ignored by MATLAB,
  but acted on and checked by LabMate. We outline the design
  principles behind LabMate, our current progress, and our future
  plans.
\end{abstract}

@@

To rectify this situation, we are developing LabMate, which is a tool
to reify current virtuous engineering practices as a formal language
of MATLAB comments. MATLAB remains in the dark, but LabMate reads,
assesses, and transforms MATLAB programs in accordance with these
comments.
%
Behind the scenes, LabMate is retrofitting an expressive type system
to MATLAB, with more of the meaning of programs recorded in their
types.
%
This gives a lightweight and low-cost way for MATLAB programmers to
express their intent, and in a language they are already working in,
rather than starting over from scratch.
%
While type systems and their theory is a well established discipline
of computer science, the development of LabMate has several novel
advances on the algebraic structure required to classify
matrices and the meanings of the quantities therein, e.g., their
physical dimensions. LabMate brings advanced type systems to
effectiveness within, rather than instead of, existing scientific and
engineering toolchains and practices.

 LabMate is under constant development; we encourage interested readers to get in touch for access to the latest version.
 At time of writing, the necessary core type system is in place: we are busily extending the variety of MATLAB expressions it can make sense of, and the constraints it can solve.
%
% LabMate is under constant development; the necessary core type system is in place: we are extending the variety of MATLAB expressions it can make sense of, and the constraints it can solve.
We are close to covering the examples in this paper, but not quite there yet.
%
We aim to eliminate the speculative future tense from our examples before camera-ready copy is required.
%
This paper effectively documents how far we have come, and the roadmap to our next milestone.

@@

An electric current with a known value (the applied signal) is passed through the resistor being measured, and a series of voltage readings are taken at fixed time intervals (the recorded output). After a specified number of readings are taken, the direction of the input signal must be reversed in order to separate the offset and drift from the voltage readings. Therefore, the applied signal (bridge energisation) is a square wave of {\em reversals}. A calibration factor is also applied, dividing the signal into two parts with different amplitudes.

The recorded output follows the input signal, superimposed on a detector with noise, offset and drift. Figure~1 %\ref{fig:rec_out}
provides an example of recorded output, using simulated data.
\begin{figure}[H]
  \begin{center}
    \includegraphics[width=0.6\linewidth]{example_output.jpg}
    \captionwide{Simulated example of recorded output.}
    \label{fig:rec_out}
  \end{center}
\end{figure}

@@

We begin by declaring the types of the input data: the \texttt{times} and the \texttt{voltages}. Our example calculation uses twelve time and voltage inputs, arranged as two rows to be read from a file. A real measurement would involve considerably larger datasets.
\begin{verbatim}
%> times    :: [ 1 x 12 ] double
%> voltages :: [ 1 x 12 ] double
\end{verbatim}
Lines starting with `\texttt{\%>}' are interpreted as ordinary comments by MATLAB, but as \remph{directives} by LabMate, here giving us enough information not only to insist that the values of \texttt{times} and \texttt{voltages} \remph{conform} to the given sizes, but to determine \remph{how} to read them from a file. We would issue the further directive
\begin{verbatim}
%> readfrom "inputs.txt" times voltages
\end{verbatim}
LabMate is a program \remph{transducer}: its output is a modified version of its input, allowing it to respond to directives with information in comments or by generating code. Here we would expect to see something like
\begin{verbatim}
%> readfrom "inputs.txt" times voltages
%<{
h=fopen("inputs.txt");
c=textscan(h,'%f');
fclose(h);
src = c{1};
readPtr = 1;
for i = 1:12
  times(i) = src(readPtr);
  readPtr = readPtr + 1;
end
for i = 1:12
  voltages(i) = src(readPtr);
  readPtr = readPtr + 1;
end
%<}
\end{verbatim}
where the comment lines `\pct\res\lcb' and `\pct\res\rcb' visually delimit the extent of the code generated by LabMate, so that humans may safely ignore it. LabMate knows enough about scope in MATLAB to ensure that auxiliary variable naming never creates conflict. We have kept to one datum per line for simplicity, but we are free to extend our format specification language as we see fit~\cite{mgen}.

@@

Under the hood, LabMate translates selected well behaved MATLAB expressions
to typed expressions in its own core theory of types and values.
%
As matrices feature heavily in MATLAB code, the types of matrices
play a central role in this theory.
%
In particular, we acknowledge that the rows and columns of a matrix might pertain to distinct individual entities, rather as the rows and columns of a spreadsheet do.
%
Internally, a matrix type looks like
\[\Matrix{R}{C}{E}{\rs}{\cs}
\]
%
We thus \param{} matrix types by two arbitrary `header' types $R$, for the individuals the rows concern, and $C$ for the columns. We may then give two lists, $\rs = [r_1,\ldots,r_m]\in\List{R}$ (so that each $r_i\in R$) and $\cs = [c_1,\ldots,c_n]\in\List{C}$ of those individuals. A \param{}d type $E(r,c)$ gives the type of entry for any combination of row and column individual, with $E(r_i,c_j)$ being the type of the entry $e_{i,j}$. In the diagram below, the $e$s inside the rectangle are the matrix entries computed by the MATLAB code. The $r$s and $c$s outside the rectangle are \remph{meta}data coming from the matrix type, known only to LabMate, and not available as data to MATLAB.
\[\begin{array}{@{}r|ccccc|}
\; & c_1      & \cdots & c_j     & \cdots & c_n     \\
\hline
r_1    & e_{1,1}  &        & e_{1,j} &        & e_{1,n} \\
\vdots &          &        &         &        &         \\
r_i    & e_{i,1}  &        & e_{i,j} &        & e_{i,n} \\
\vdots &          &        &         &        &         \\
r_m    & e_{m,1}  &        & e_{m,j} &        & e_{m,n}  \\
\hline
\end{array}\]
To check that $\Matrix{R}{C}{E}{\rs}{\cs}$ is a valid type, LabMate must:
\begin{enumerate}
\item check that $R$ and $C$ are both valid types;
\item check that $E(r,c)$ is a valid type whenever $r\in R$ and $c\in C$;
\item check that $\rs\in\List{R}$ and $\cs\in\List{C}$.
\end{enumerate}

@@

LabMate elaborates MATLAB expressions and commands into terms of its own internal core type theory.
%
Although the simple examples in this paper use \emph{constant} sizes and dimensions, any practicable library requires size, header and dimension expressions to contain \emph{parameters}.
%
For example, if \texttt{A} is $i\times j$ and \texttt{B} is $i\times k$, then
\begin{verbatim}
[ A B
  B A ]
\end{verbatim}
should be a valid matrix pasting, because $j+k = k+j$. Our notion of equality for numbers thus includes tacit applications of commutative and associative laws. % induces commutative monoid structure.
%
Physical dimensions are modelled as the free Abelian group generated by some base set of dimensions~\cite{dim.group}, so we decide that theory, too.
%
The core theory we need is in place.

@@

Coming back to the example from Section~\ref{sec:example} again, we aim to give both \texttt{M} and \texttt{voltages} more precise types, also taking their physical dimensions into account,
\begin{verbatim}
%> M :: [ 12 x k::D<-[{},{},{},{},T] ] Q(k)
%> voltages :: [ 1 x 12 ] Q(V)
\end{verbatim}
  but we had better unpack what we mean.

@@

Secondly, we see \texttt{Q(}$d$\texttt{)} as the type of \remph{quantities} with physical dimension $d$, represented in MATLAB as a mere \texttt{double} with standardised units. We shall define a type $\texttt{D}$ of dimensions presently. The notation
\begin{verbatim}
k::D<-[{},{},{},{},T]
\end{verbatim}
will tell LabMate that the column header type is $\texttt{D}$, and that the matrix entry type \texttt{Q(k)} varies in accordance with the header, locally named \texttt{k}, which is drawn from the given list of dimensions.
%
But what type is \texttt{D}? We will have defined, in LabMate directives, rather than MATLAB,
\begin{verbatim}
%{>
Base =
  ['Time, 'Length, 'Mass, 'Current,
   'Temperature, 'Amount, 'Luminosity]
D = Abel(Enum(Base))
%}
\end{verbatim}
where \texttt{Base} is a list of symbolic constants standing for base dimensions, \texttt{Enum} converts any such list to a finite enumeration type, and \texttt{Abel} constructs the free Abelian group on any type of generators, with unit \lcb\rcb, singleton generators \lcb$g$\rcb, multiplication \texttt{*}, division \texttt{/}, and exponentiation by integer constants \crt. We further define
\begin{verbatim}
%{>
T = {'Time}
V = {'Mass}*{'Length}^2/T^3/{'Current}
%}
\end{verbatim}
and \texttt{voltages} becomes a row vector uniformly containing quantities in \texttt{Q(V)}.

With this refinement in place
\begin{verbatim}
x = M \ voltages'
%> typeof x
\end{verbatim}
will yield
\begin{verbatim}
%< x :: [ k::D<-[{},{},{},{},T] ] Q(V/k)
\end{verbatim}

@@

However, if all LabMate does is complain about mistakes, it will not appeal to programmers in a hurry.
%
We have already shown a simple example where LabMate does not merely check but rather generates code using its types as specification.
%
There is considerable potential to push further in this direction, little by little.
%
For example, we could equip LabMate with its own expression language for calculating with quantities which have not only dimensions but \remph{units}, compiling to MATLAB with appropriate conversion factors correctly inserted.
%
Our mission is, ever so gently, to tilt the balance towards programming at a higher level of structure and meaning in LabMate's directive language.

@@

Inspired again by Kennedy's work --- his gradualist reformation of PHP~\cite{hack} ---
we have set out on this journey to bring the benefits of expressive type systems to scientists and engineers \remph{where they are}.
%
It would be unrealistic of us to expect practitioners to abandon their existing toolchains and learn whole new programming styles and languages that, in any case, exist only as academic research prototypes.
%
Moreover, it has very much \remph{not} turned out that, in addressing MATLAB, we were merely \remph{applying} type system research already well established.
%
This project has pushed us in new directions and generated raised expectations of the \remph{algebra} to be done by the typechecker.
%
There is so much to learn from applications of dependent types in the wild!

@@

Thanks to NPL colleague Nick Fletcher for his help and guidance providing the example from Section \ref{sec:example}. Thanks also to NPL colleagues Peter Harris, Louise Wright and Ian Smith for reviewing this paper, and to Professor Neil Ghani for his support.

@@

% LocalWords:  Idris
